
---
title: Assignment 1
author: "Osaro Orebor 1168827, Calvin Boateng, Jair"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: united
---


```{r setup1, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Our Research question is: Which audio features best predict a song’s popularity on Spotify?

We aimed to explore whether certain musical characteristics—such as danceability, loudness, speechiness, and year of release—have a significant effect on a song's popularity. We also compared different modeling approaches (linear regression and Lasso regression) to identify which predictors are most useful and how well these models perform.


# Setting Libraries

```{r Libraries2, message = FALSE, warning = FALSE}
suppressWarnings({
  library(dplyr)
  library(ggplot2)
  library(ggpubr)
  library(kableExtra)
  library(tidyverse)
  library(readr)
  library(knitr)
  library(weathermetrics)
  library(gridExtra)
  library(plotly)
  library(GGally)
  library(magrittr)
  library(regclass)
  library(MASS)
  library(pROC)
  library(caret)
  library(car)
})
```

```{r set path}
# Set seed for reproducibility
set.seed(123)

```
# Loading in the Data


```{r loading in the file, echo=FALSE}
data <- read.csv("data/Spotify-2000.csv")



```
.

# Explorative Data Analysis 

## looking at the data and analyse the variables used


### Summary
```{r Data summary, echo = FALSE}
kable(head(data, 10)) %>% 
  kable_styling("striped", full_width = F) %>% 
  scroll_box(width = "800px", height = "500px")

kable(tail(data, 10)) %>% 
  kable_styling("striped", full_width = F) %>% 
  scroll_box(width = "800px", height = "500px")

str(data)
```

##Mutating Data

``` {r}
#mutate our data so character values become factors
data <- data %>%
  mutate_if(is.character, as.factor) %>%
  mutate(Length..Duration. = as.integer(Length..Duration.))
str(data)
```

# Plots

## Plot 1
``` {r Heatmap}
numeric_columns <- data %>%
  dplyr::select(where(is.numeric)) %>%
  dplyr::select(-any_of("ID"))

# Plot de heatmap 
ggcorr(
  numeric_columns, 
  label = FALSE,            
  layout.exp = 1,
  low = "#7FFF7F",          # groen voor lage correlatie
  mid = "white", 
  high = "#FF4C4C",         # rood voor hoge correlatie
  name = "Correlatie"
) +
  labs(
    title = "Correlatie tussen numerieke variabelen",
    caption = "Rood = sterke correlatie, Groen = zwak of geen correlatie"
  )

```

## Plot 2
``` {r}
data_long <- data %>%
  dplyr::select(Year, Danceability, Energy, Loudness..dB., Popularity) %>%
  pivot_longer(cols = c(Danceability, Energy, Loudness..dB., Popularity),
               names_to = "Feature",
               values_to = "Value") %>%
  group_by(Year, Feature) %>%
  summarise(Avg_Value = mean(Value, na.rm = TRUE), .groups = "drop")

ggplot(data_long, aes(x = Year, y = Avg_Value, color = Feature)) +
  geom_line(size = 1) +
  labs(title = "Trends of some musical features and popularity over time",
       x = "Year",
       y = "Average Value") +
  theme_minimal()
```

## Plot 3 

# Regression

## Lasso-Regression

# Conclusion

# Contribution
Osaro:
Calvin:
Jair:
