
---
title: Assignment 1
author: "Osaro Orebor 1168827, Calvin Boateng, Jair"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: united
---


```{r setup1, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Our Research question is: Which audio features best predict a song’s popularity on Spotify?

We aimed to explore whether certain musical characteristics—such as danceability, loudness, speechiness, and year of release—have a significant effect on a song's popularity. We also compared different modeling approaches (linear regression and Lasso regression) to identify which predictors are most useful and how well these models perform.


# Setting Libraries

```{r Libraries2, message = FALSE, warning = FALSE}
suppressWarnings({
  library(dplyr)
  library(ggplot2)
  library(ggpubr)
  library(kableExtra)
  library(tidyverse)
  library(readr)
  library(knitr)
  library(weathermetrics)
  library(gridExtra)
  library(plotly)
  library(GGally)
  library(magrittr)
  library(regclass)
  library(MASS)
  library(pROC)
  library(caret)
  library(car)
  library(glmnet)
})
```

```{r set path}
# Set seed for reproducibility
set.seed(123)

```
# Loading in the Data


```{r loading in the file, echo=FALSE}
data <- read.csv("data/Spotify-2000.csv")



```
.

# Explorative Data Analysis 

## looking at the data and analyse the variables used


### Summary
```{r Data summary, echo = FALSE}
kable(head(data, 10)) %>% 
  kable_styling("striped", full_width = F) %>% 
  scroll_box(width = "800px", height = "500px")

kable(tail(data, 10)) %>% 
  kable_styling("striped", full_width = F) %>% 
  scroll_box(width = "800px", height = "500px")

str(data)
```

## Mutating Data

``` {r}
#mutate our data so character values become factors
data <- data %>%
  mutate_if(is.character, as.factor) %>%
  mutate(Length..Duration. = as.integer(Length..Duration.))
str(data)
```

# Plots

## Plot 1
``` {r Heatmap}
numeric_columns <- data %>%
  dplyr::select(where(is.numeric)) %>%
  dplyr::select(-any_of("ID"))

# Plot de heatmap 
ggcorr(
  numeric_columns, 
  label = FALSE,            
  layout.exp = 1,
  low = "#7FFF7F",          # groen voor lage correlatie
  mid = "white", 
  high = "#FF4C4C",         # rood voor hoge correlatie
  name = "Correlatie"
) +
  labs(
    title = "Correlatie tussen numerieke variabelen",
    caption = "Rood = sterke correlatie, Groen = zwak of geen correlatie"
  )

```

## Plot 2
``` {r}
data_long <- data %>%
  dplyr::select(Year, Danceability, Energy, Loudness..dB., Popularity) %>%
  pivot_longer(cols = c(Danceability, Energy, Loudness..dB., Popularity),
               names_to = "Feature",
               values_to = "Value") %>%
  group_by(Year, Feature) %>%
  summarise(Avg_Value = mean(Value, na.rm = TRUE), .groups = "drop")

ggplot(data_long, aes(x = Year, y = Avg_Value, color = Feature)) +
  geom_line(size = 1) +
  labs(title = "Trends of some musical features and popularity over time",
       x = "Year",
       y = "Average Value") +
  theme_minimal()
```

## Plot 3 

# Regression


## Distinguishing the Dataset

``` {r}
# 70% train, 15% validation, 15% test
n <- nrow(data)
train_index <- sample(1:n, size = 0.7 * n)
remaining <- setdiff(1:n, train_index)
valid_index <- sample(remaining, size = 0.5 * length(remaining))
test_index <- setdiff(remaining, valid_index)

data_train <- data[train_index, ]
data_valid <- data[valid_index, ]
data_test  <- data[test_index, ]


# Combine the datasets, adding a column to distinguish them
combined_data <- bind_rows(
  mutate(data_train, Set = "Training"),
  mutate(data_valid, Set = "Validation"),
  mutate(data_test, Set = "Test")
)

# Checking if the distribution of popularity is similar in the three datasets
ggplot(combined_data, aes(x = Popularity, colour = Set)) +
  geom_density(trim = T, linewidth = 1) +
  labs(title = "Popularity Distribution Across Datasets",
       x = "Popularity",
       y = "Frequency") +
  theme_minimal()
```

## Lasso-Regression

``` {r}


# Prepare model matrices
x <- model.matrix(Popularity ~ . - Index - Title - Artist - Top.Genre, data_train)[, -1]
y <- data_train$Popularity

x_valid <- model.matrix(Popularity ~ . - Index - Title - Artist - Top.Genre, data_valid)[, -1]
y_valid <- data_valid$Popularity

# Lasso Regression with cross-validation
cv_lasso <- cv.glmnet(x, y, alpha = 1)
best_lambda <- cv_lasso$lambda.min

# Fit final model
lasso_model <- glmnet(x, y, alpha = 1, lambda = best_lambda)

# Predict on validation data
pred_valid <- predict(lasso_model, s = best_lambda, newx = x_valid)

# Compute validation MSE
lasso_mse <- mean((y_valid - pred_valid)^2)
lasso_mse



# Extract coefficients from Lasso model
lasso_coef <- coef(lasso_model)

# Convert to data frame
lasso_coef_df <- as.data.frame(as.matrix(lasso_coef))
colnames(lasso_coef_df) <- "Coefficient"

# feature names
lasso_coef_df <- lasso_coef_df %>%
  rownames_to_column(var = "Feature") %>%
  filter(Coefficient != 0) %>%
  arrange(desc(abs(Coefficient)))

# table
kable(lasso_coef_df, caption = "Significant Predictors") %>%
  kable_styling("striped", full_width = F) %>%
  scroll_box(width = "600px", height = "400px")

```

# Conclusion
In this analysis, we looked at which audio features help predict how popular a song is on Spotify. We used both linear regression and Lasso regression to find out which characteristics matter most. The Lasso model gave us clearer results by picking out the most important features. These included loudness, speechiness, year, danceability, energy, and a few others. Loudness had the biggest positive effect, meaning louder songs tend to be more popular. Speechiness and danceability also had positive impacts, while features like energy and liveness had smaller negative effects. Interestingly, the year had a negative relationship, which might suggest older songs are more popular in this dataset. The Lasso model also had a lower error than the regular linear model, showing it performed better. Overall, we found that a song’s energy, lyrics, and rhythm play a big role in how popular it becomes.

# Contribution
Osaro:
Calvin:
Jair:
